---
title: "Themes in Mark Carney's First Year"
subtitle: "A text analysis of official Prime Minister of Canada speeches"
author: "Daniel Sass"
date: "2026-06-16"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    include-in-header:
      text: |
        <style>
        @media (min-width: 992px) {
          #quarto-margin-sidebar {
            position: sticky;
            top: 1rem;
            align-self: start;
            max-height: calc(100vh - 2rem);
          }

          #TOC {
            position: relative;
            max-height: calc(100vh - 2rem);
            overflow-y: auto;
            padding-left: 0.85rem;
          }

          #TOC ul.collapse {
            display: block;
            height: auto;
            visibility: visible;
          }

          #TOC::before {
            content: "";
            position: absolute;
            top: 2.5rem;
            bottom: 0;
            left: 0;
            width: 1px;
            background: rgba(0, 0, 0, 0.16);
          }

          #TOC .toc-scroll-marker {
            position: absolute;
            top: 0;
            left: 0;
            width: 3px;
            height: 1.45rem;
            border-radius: 999px;
            background: #3B6EA8;
            transform: translateY(var(--toc-marker-top, 2.5rem));
            transition: transform 140ms ease;
          }

          #TOC .nav-link.active {
            color: #3B6EA8;
            font-weight: 600;
          }
        }
        </style>
        <script>
        document.addEventListener("DOMContentLoaded", function () {
          const toc = document.querySelector("#TOC");
          if (!toc) return;

          const links = Array.from(toc.querySelectorAll("a.nav-link[data-scroll-target]"));
          const sections = links
            .map(function (link) {
              return {
                link: link,
                section: document.querySelector(link.getAttribute("data-scroll-target"))
              };
            })
            .filter(function (item) {
              return item.section;
            });

          if (!sections.length) return;

          const marker = document.createElement("div");
          marker.className = "toc-scroll-marker";
          marker.setAttribute("aria-hidden", "true");
          toc.prepend(marker);

          let ticking = false;

          function setActive(link) {
            links.forEach(function (item) {
              item.classList.toggle("active", item === link);
            });

            const tocRect = toc.getBoundingClientRect();
            const linkRect = link.getBoundingClientRect();
            const markerTop = toc.scrollTop + linkRect.top - tocRect.top + (linkRect.height / 2) - (marker.offsetHeight / 2);
            marker.style.setProperty("--toc-marker-top", markerTop + "px");
          }

          function updateTocMarker() {
            const offset = 140;
            let active = sections[0].link;

            sections.forEach(function (item) {
              if (item.section.getBoundingClientRect().top <= offset) {
                active = item.link;
              }
            });

            setActive(active);
            ticking = false;
          }

          function requestUpdate() {
            if (!ticking) {
              window.requestAnimationFrame(updateTocMarker);
              ticking = true;
            }
          }

          window.addEventListener("scroll", requestUpdate, { passive: true });
          window.addEventListener("resize", requestUpdate);
          toc.addEventListener("scroll", requestUpdate, { passive: true });
          updateTocMarker();
        });
        </script>
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(tidytext)
library(topicmodels)
library(scales)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

```{r}
speeches <- read_csv("Data/carney_pm_speeches_raw.csv", show_col_types = FALSE) |>
  mutate(
    date = as.Date(date),
    doc_id = row_number(),
    title = if_else(is.na(title) | title == "", "Untitled speech", title)
  )

custom_stop_words <- tibble(word = c("canada", "build"))

clean_tokens <- speeches |>
  select(doc_id, date, title, text) |>
  unnest_tokens(word, text) |>
  mutate(word = str_replace(word, "’s$|'s$", "")) |>
  anti_join(stop_words, by = "word") |>
  mutate(word = case_when(
    word %in% c("canadian", "canadians") ~ "canada",
    word %in% c("building", "built") ~ "build",
    word %in% c("investments") ~ "investment",
    word %in% c("partnerships", "partners") ~ "partnership",
    word %in% c("projects") ~ "project",
    word %in% c("stronger", "strength") ~ "strong",
    word %in% c("nations", "national") ~ "nation",
    word %in% c("economic") ~ "economy",
    word %in% c("creating") ~ "create",
    word %in% c("homes", "housing") ~ "home",
    TRUE ~ word
  )) |>
  anti_join(custom_stop_words, by = "word")

word_counts_by_doc <- clean_tokens |>
  count(doc_id, word, sort = TRUE)

speech_dtm <- word_counts_by_doc |>
  cast_dtm(document = doc_id, term = word, value = n)

carney_lda <- LDA(speech_dtm, k = 4, control = list(seed = 712))

topic_terms <- tidy(carney_lda, matrix = "beta")
topic_documents <- tidy(carney_lda, matrix = "gamma") |>
  mutate(document = as.integer(document)) |>
  left_join(speeches |> select(doc_id, date, title), by = c("document" = "doc_id"))

top_topic_terms <- topic_terms |>
  group_by(topic) |>
  slice_max(beta, n = 10, with_ties = FALSE) |>
  ungroup() |>
  arrange(topic, desc(beta))

topic_labels <- tibble(
  topic = 1:4,
  topic_name = c(
    "Housing, Infrastructure, and Economy",
    "National Security and Sovereignty",
    "Defence, Trade, and Global Partnerships",
    "AI, Communities, and Major Projects"
  )
)

topic_terms_named <- top_topic_terms |>
  left_join(topic_labels, by = "topic") |>
  mutate(topic_label = str_c("Topic ", topic, ": ", topic_name))

topic_documents_named <- topic_documents |>
  left_join(topic_labels, by = "topic") |>
  mutate(topic_label = str_c("Topic ", topic, ": ", topic_name))
```

## Motivation

This project is a recap of Mark Carney's first year in office through the language of his official speeches. Rather than evaluate policy outcomes directly, the goal is to identify the themes that appeared most often in his public remarks and to make those patterns visible in a reproducible way.

The guiding question is:

> What core themes defined Mark Carney's official speeches during the first year covered by this corpus?

## Source

The analysis uses speeches scraped from the official Prime Minister of Canada website and saved in `Data/carney_pm_speeches_raw.csv`.

```{r}
corpus_summary <- speeches |>
  summarise(
    `Speeches analyzed` = n(),
    `First speech date` = min(date),
    `Last speech date` = max(date),
    `Missing speech texts` = sum(is.na(text) | text == "")
  )

kable(corpus_summary)
```

```{r}
speeches |>
  count(month = floor_date(date, "month")) |>
  ggplot(aes(month, n)) +
  geom_col(fill = "#3B6EA8", width = 25) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  labs(
    title = "Speech coverage by month",
    x = NULL,
    y = "Number of speeches"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Methods

I used a lightweight text-analysis workflow in R:

1. Scraped official speech pages from the Prime Minister of Canada website.
2. Tokenized each speech into individual words.
3. Removed common English stop words.
4. Grouped a small set of related word forms, such as `investments` into `investment` and `projects` into `project`.
5. Removed `canada` and `build` after normalization because they dominated every topic and made the model less informative.
6. Estimated a four-topic Latent Dirichlet Allocation model.
7. Suggested plain-language topic names by reading the highest-probability terms in each topic.

This is an exploratory topic model, not a definitive classification of every speech.

## Key Visuals

### Most Frequent Terms

```{r}
clean_tokens |>
  count(word, sort = TRUE) |>
  slice_max(n, n = 20) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word)) +
  geom_col(fill = "#2F7D5C") +
  labs(
    title = "Most frequent terms after cleaning",
    subtitle = "Common stop words plus the generic terms 'canada' and 'build' are removed",
    x = "Frequency",
    y = NULL
  )
```

### Suggested Topic Names

```{r}
topic_table <- topic_labels |>
  left_join(
    top_topic_terms |>
      group_by(topic) |>
      summarise(`Top terms` = str_c(term, collapse = ", "), .groups = "drop"),
    by = "topic"
  ) |>
  rename(`Topic` = topic, `Suggested name` = topic_name)

kable(topic_table)
```

```{r}
topic_terms_named |>
  mutate(term = reorder_within(term, beta, topic_label)) |>
  ggplot(aes(beta, term, fill = topic_label)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic_label, scales = "free_y") +
  scale_y_reordered() +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Top words by topic",
    x = "Estimated probability within topic",
    y = NULL
  )
```

### Topic Emphasis Over Time

```{r}
topic_documents_named |>
  group_by(month = floor_date(date, "month"), topic_label) |>
  summarise(mean_gamma = mean(gamma), .groups = "drop") |>
  ggplot(aes(month, mean_gamma, color = topic_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Average topic emphasis by month",
    x = NULL,
    y = "Average topic share",
    color = NULL
  ) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "bottom"
  )
```

## Main Findings

Four broad themes stand out after removing generic terms that appeared across nearly all speeches:

- **Housing, infrastructure, and economy:** Frequent terms such as `home`, `project`, `economy`, `investment`, and `energy` point to a strong focus on domestic economic development and large-scale public priorities.
- **National security and sovereignty:** Terms including `nation`, `strong`, `country`, `security`, `forces`, and `protect` suggest a repeated emphasis on state capacity, protection, and sovereignty.
- **Defence, trade, and global partnerships:** The model separates a topic around `world`, `defence`, `partnership`, `trade`, `security`, `strategic`, and `global`, reflecting international positioning and alliance-building.
- **AI, communities, and major projects:** Terms such as `ai`, `communities`, `indigenous`, `community`, `project`, and `government` capture a mix of technology strategy, public investment, and community-facing commitments.

A broader pattern also appears: Carney's speeches often connect domestic prosperity with external security. Economic language and security language are not isolated; they frequently appear as part of the same governing narrative.

## Limitations

- The corpus only includes official speeches from the Prime Minister of Canada website. It does not include interviews, social media, parliamentary debate, cabinet statements, or media coverage.
- Topic names are interpretive labels based on top model terms. Another reader might choose slightly different labels.
- The choice of four topics is a modeling decision, not a natural fact about the speeches.
- Longer speeches can contribute more text and may have more influence on the model.
- The analysis identifies recurring language, not whether policies succeeded or whether the rhetoric was accurate.
- The scraping logic depends on the structure of the source website at the time the data was collected.

## Reproducibility

The public-facing report uses the saved CSV in `Data/carney_pm_speeches_raw.csv`, so it can be rendered without re-scraping the website.

To reproduce the report:

```bash
quarto render carney_first_year_report.qmd
```

The technical notebook, `NLP_draft1.qmd`, contains the earlier scraping and modeling proof-of-work.
